The benefits of reporting critical effect size values

### 2024-12-14 paper/paperDL.Rmd
^^

    \renewcommand{\subparagraph}[1]{\@startsection{subparagraph}{5}{1em}%
      {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
      {-\z@\relax}%
      {\normalfont\normalsize\bfseries\itshape\hspace{\parindent}{#1}\textit{\addperi}}{\relax}}
    \makeatother
    \usepackage{xcolor}
---

@@

The "critical effect size" refers to the smallest effect size that can be statistically significant, given the test performed, the sample size, and the alpha level [@lakens2022sample]. Consider a study that tests a bivariate correlation with $n = 20$ participants and uses $\alpha < 0.05$ as the threshold for statistical significance in a two-sided test. The critical effect size values for this test are Pearson's $r = -0.44$ and $r = 0.44$, which means that the statistical test will yield a significant result for effects smaller than $r = -0.44$ or larger than $r = 0.44$, while observed effect sizes between these two values will lead to a statistically non-significant result. In the present paper, we suggest that researchers should systematically report critical effect size values when they make inference use significance levels in hypothesis tests to make claims about the presence or absence of effects. Critical effect size values can serve as a useful complement interpret study results, communicate the informational value of a study design, and contextualizes the interporetation of non-signiicant results, particularly when statistical significance is relied upon for inference. Reporting critical effect size values is especially important when the sample size was not (or could not) be predetermined on the basis of a smallest effect size of interest.

Null hypothesis significance testing (NHST) remains the most prominent approach for statistical inference in science, even though there are widespread concerns about the misuse of hypothesis tests [@cohen1994; @gigerenzer2004null; @krueger2001null]. Over the past decades, numerous proposals have emerged to improve the use of hypothesis testing. These include complementing hypothesis tests with effect sizes and their confidence intervals, preregistering hypotheses before the data is collected, and conducting power analyses to determine the sample size \textit{a priori} based on a smallest effect size of interest. Researchers often publish studies with low power for medium to small effect sizes [@szucs2017empirical], with the main justification for the sample size being resource limitations [@lakens2022sample]. However, low power makes it challenging to distinguish signal from noise, and combined with the selective reporting of statistically significant results in the published literature leads to an overestimation of effect sizes [@altoe2020enhancing]. Certain types of research questions can be studied by relying on online data collection, which has made it cheaper and more feasible to collect very large samples. Studies with very large samples are well-powered to detect even small effect sizes, but they also require researchers to carefully consider the possibility that an effect might be statistically significant, but practically insignificant.

We believe there are two clear benefits to reporting critical effect size values for a corresponding test. First, when sample sizes are small, the critical effect size values inform readers about whether the effect sizes that could lead to rejecting the null hypothesis are in line with realistic expectations. If the sample size is small and only very large effects would yield a statistically significant result, and the underlying mechanism that is examined is unlikely to lead to such large effect sizes, researchers will realize they are not able to collect sufficient data to perform an informative hypothesis test. An a-priori power analysis would typically lead to a similar conclusion, but reporting critical effect size values will focus the attention more strongly on which effect sizes are reasonable to expect. Second, when sample sizes are very large, the critical effect size values will make it clear that trivially small effect sizes will be statistically significant. This will focus the attention of researchers on the difference between statistical significance and practical significance, and raises awareness of the importance to interpret the size of the effect. This is especially important in correlational studies with large sample sizes in psychology, where systematic but uncontrolled sources of variability may lead to small but theoretically meaningless non-zero effects, a phenomenon referred to as the 'crud factor' [@orben_crud_2020].

Let us consider the following two scenarios as examples. First, imagine researchers who conduct a study involving a between-group comparison. Due to severe resource constraints, they are only able to collect a limited sample size of $n = 30$ ($n = 15$ per group). The authors did not perform an a-priori power analysis, and their sample size justification is purely based on how many participants they can collect. The statistical tests yields a non-significant result, $p > 0.05$. Given their prior expectation that the effect of interest may not be large, they acknowledge that their study was likely underpowered, though without a theoretically expected effect size, it remains uncertain how low the statistical power is. Subsequently, they compute the critical effect size values, revealing Cohen's $d = -0.75$ and $d = 0.75$. This indicates that any observed Cohen's $d$ between -0.75 and 0.75 will certainly fail to reach statistical significance. By reporting these critical effect size values the researchers transparently convey that estimated effects much larger than most effects in the psychological literature will always fall short of significance. This means that the hypothesis test will not be informative. Non-significant effects are likely, even if there is a true effect of interest. If a significant effect is observed, but most effects in a field are smaller than a study can detect, it is probable the effect size is overestimated  [@hedges1984estimation], and this should be taken into account when interpreting the effect size.

In a second scenario, imagine researchers who gain access to a very large archival dataset ($n = 5,000$) and decide to explore bivariate correlations bwetween variables. With such a large sample size, very small correlations will reach statistical significance, and the critical effect size values for a two-sided test are $r = \pm 0.03$. It is worth noting that such small effects may potentially reflect just minor artifacts [@wilson2020science], such as subtle experimenter effects or slight non-independence among observations, even in meticulously designed studies. A similar scenario may arise in meta-analysis. Despite potential loss of precision due to substantial heterogeneity across effect sizes in different studies, meta-analyses typically synthesize a large amount of evidence, and as a consequence, even very small average effect sizes can reach significance. While the focus of meta-analysis is generally on estimating effect sizes with uncertainty, statistical significance is routinely reported and interpreted. Signaling the critical effect size values beforehand can serve as a clear warning that statistically significant results should not automatically be interpreted as practically significant, thus urging caution when interpreting the results.

To provide more tangible illustrations of the practical applications of critical effect size values, let us examine two real-world instances drawn from published research. The first one features a small sample and has been targeted in the series of replications by the Open Science Collaboration [@open2015estimating], while the second study had a very large sample, and has attracted attention of the scientific community. The first case pertains to "Study 5" conducted by @mccrea2008self. This experiment compared two groups based on their performance percentage in a math test following a preliminary practice session with feedback of failure. One group was exposed to failure-excusing, self-handicapping thoughts, while both groups underwent the math test afterward. With a modest sample size of 28 participants split into two groups (13 and 15 participants, respectively), a directional one-tailed t-test was conducted on the percentage of correct answers to compare the groups. The resulting effect was statistically significant, $t(26) = 1.87$, $p$ (one-tailed) $< .05$. Based on the reported data, Cohen's $d = 0.736$, the findings indicated a considerable effect size. However, the critical value for Cohen's $d$ in this context was $0.645$, suggesting that, given this sample size, statistical significance is achieved only for estimated effects of substantial magnitude. As the study was not a Registered Report, selective reporting of statistically significant findings was common at the time of publication, and most effects in psychology are substantially smaller, the effect size estimate is likely upwardly biased. Subsequent replication efforts [@open2015estimating] reaffirmed the significance of the effect, and used a somewhat larger sample ($n = 61$), $t(59)=2.325$. The estimated effect size was somehow lower than that originally reported, however, $d = 0.605$ (here, critical d = 0.428). THIS IS NOT A GOOD EXAMPLE, BECAUSE THESE TWO EFFECT SIZE ESTIMATES DO NOT DIFFER STATISTICALLY FROM EACH OTHER, SO WE CAN NOT SAY THE TWO EFFECT SIZES DIFFER! THERE SHOULD BE OTHER EXAMPLES IN THE LITERATURE, MAYBE FROM REGISTERED REPLICATION REPORTS.

The second scenario involves the study by @kramer2014experimental, which explored the impact of emotional content on Facebook users' experiences. With a notably large sample size of $n = 689.003$, the researchers observed a statistically significant increase in the number of negative words typed by participants when positive posts were reduced in their timeline, $t(310,044) = −5.63$, $p < 0.001$ (non-directional test). Based on this and other results from the study, the authors conclude that emotions expressed by other users on Facebook influence our own emotions [CHANGE A BIT, STAY CLOSER TO WHAT THE RESEARCHERS SAY, MAYBE QUOTE THEM DIRECTLY]. The critical values for Cohen's $d$ in this experimental design are $-0.006$ and $+0.006$. It should have been clear from the outset that the researchers should have considered the effect size beyond merely testing for the statistical significancee. The critical effect size should prompt reflection on which effect sizes are truly meaningful in this context, and whether the observed effect size ($d = 0.02$) is theoretically or practically interesting.

@@

In this section, we provide guidance and formulas for computing critical effect size values in general [BUT WE FOCUS ON STANDARDIZED, NOT RAW, EFFECT SIZES - SO WE EITHER GIVE A REASON FOR THIS, OR WE ALSO ADD RAW EFFECTS], with examples for frequently encountered effect sizes including Standardized Mean Differences (Cohen's $d$), correlations (Pearson's $r$), and raw and standardized coefficients in linear models. These formulas have been incorporated into R functions of the package `criticalESvalue` accessible at: [https://github.com/psicostat/criticalESvalue](https://github.com/psicostat/criticalESvalue), and are elaborated upon in the subsequent section.

@@

Where $b$ is the unstandardized effect size that depends on the type of test. [THE NEXT SENTENCE NEEDS TO BE REWRITTEN] For example, in the two sample $t$-test or the single mean for a one sample $t$-test. The denominator is the standard error of the numerator that depends on the sample size and the sample(s) variance.

Similarly, Equation \@ref(eq:eq-gen-d) [REWRITE THE REST OF THIS SENTENCE] formalize a general form of the effect size.

@@

Where $b$ is the unstandardized effect size and $s$ is the standardization term. For example, in the two-sample case, $s$ is the pooled standard deviation between the two samples, while in the paired samples case, $s$ is the standard deviation of the differences for paired samples.

@@

NOT SURE TO KEEP THIS! [DL: I THINK IT IS MAYBE FOR AN APPENDIX]

>>>>

    \renewcommand{\subparagraph}[1]{\@startsection{subparagraph}{5}{1em}%
      {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
      {-\z@\relax}%
      {\normalfont\normalsize\bfseries\itshape\hspace{\parindent}{#1}\textit{\addperi}}{\relax}}
    \makeatother
    \usepackage{xcolor}
    \newcommand{\bfilippo}{\color{blue}}
    \newcommand{\efilippo}{\color{black}}
---

@@

The "critical effect size" refers to the smallest effect size that can be statistically significant, given the test performed, the sample size, and the alpha level [@lakens2022sample]. Consider a study that tests a bivariate correlation with $n = 20$ participants and uses $\alpha < 0.05$ as the threshold for statistical significance in a two-sided test. The critical effect size values for this test are Pearson's $r = -0.44$ and $r = 0.44$, which means that the statistical test will yield a significant result for effects larger than $r = -0.44$ or $r = 0.44$, while observed effect sizes between these two values will lead to a statistically non-significant result. In the present paper, we suggest that researchers should systematically report critical effect size values when they make inference use significance levels in hypothesis tests to make claims about the presence or absence of effects. Critical effect size values can serve as a useful complement for interpreting study results, communicate the informational value of a study design, and contextualize the interpretation of non-significant results, particularly when statistical significance is relied upon for inference. Reporting critical effect size values is especially important when the sample size was not (or could not) be predetermined on the basis of a smallest effect size of interest.

Null hypothesis significance testing (NHST) remains the most prominent approach for statistical inference in science, even though there are widespread concerns about the misuse of hypothesis tests [@lakens2021practical; @chow1988significance; @cortina1997logic; @hagen1997praise; @haig2017tests; @miller2017hypothesis; @nickerson2000null; @cohen1994; @gigerenzer2004null; @krueger2001null]. Over the past decades, numerous proposals have emerged to improve the use of hypothesis testing. These include complementing hypothesis tests with effect sizes and their confidence intervals, preregistering hypotheses before the data is collected, and conducting power analyses to determine the sample size \textit{a priori} based on a smallest effect size of interest. Researchers often publish studies with low power for medium to small effect sizes [@szucs2017empirical], with the main justification for the sample size being resource limitations [@lakens2022sample]. However, low power makes it challenging to distinguish signal from noise, and combined with the selective reporting of statistically significant results in the published literature leads to an overestimation of effect sizes [@altoe2020enhancing]. Certain types of research questions can be studied by relying on online data collection, which has made it cheaper and more feasible to collect very large samples. Studies with very large samples are well-powered to detect even small effect sizes, but they also require researchers to carefully consider the possibility that an effect might be statistically significant, but practically insignificant.

We believe there are two clear benefits to reporting critical effect size values for a corresponding test. In the following sections we will illustrate examples of common scenarios and we will go through the main problems and possible solutions, highlighting the utility of critical effect size values in such contexts.

## Small Sample Sizes and Uncertain sample size determination

First, when sample sizes are small, the critical effect size values inform readers about whether the effect sizes that could lead to rejecting the null hypothesis are in line with realistic expectations. If the sample size is small and only very large effects would yield a statistically significant result, and the underlying mechanism that is examined is unlikely to lead to such large effect sizes, researchers will realize they are not able to collect sufficient data to perform an informative hypothesis test. An a-priori power analysis would typically lead to a similar conclusion, but reporting critical effect size values will focus the attention more strongly on which effect sizes are reasonable to expect. \textcolor{red}{In case of already conducted studies with small sample sizes it could be argued that it would be more informative to use retrospective design analysis, but this would require both knowledge on the plausible effect size and/or on the smallest effect size of interest (SESOI), which are not always easly to determine. The use of critical effect size values can be used in a simple an efficient way to evluate which findings could have not be found significant due to sample size limitations and based on the complexity of the test.}

Let us consider the following two scenarios as examples. First, imagine researchers who conduct a study involving a between-group comparison. Due to severe resource constraints, they are only able to collect a limited sample size of $n = 30$ ($n = 15$ per group). The authors did not perform an a-priori power analysis, and their sample size justification is purely based on how many participants they can collect. The statistical tests yields a non-significant result, $p > 0.05$. Given their prior expectation that the effect of interest may not be large, they acknowledge that their study was likely underpowered, though without a theoretically expected effect size, it remains uncertain how low the statistical power is. Subsequently, they compute the critical effect size values, revealing Cohen's $d = -0.75$ and $d = 0.75$. This indicates that any observed Cohen's $d$ between -0.75 and 0.75 will certainly fail to reach statistical significance. By reporting these critical effect size values the researchers transparently convey that estimated effects much larger than most effects in the psychological literature will always fall short of significance. This means that the hypothesis test will not be informative.

\textcolor{red}{To provide a more tangible illustration of the practical applications of the critical effect sizes, let us examine two real-world instances drawn from published research. The first study was conducted on musicians and non musicians to detect differences in working memory, a modest sample size of 57 participants (42 musicians and 15 controls) was collected (@weiss2014spatial). The two groups were compared on a variety of tasks, related to verbal working memory (VWM) and tasks related to auditory skills. Regarding the VWM, the syllable-span task was administered and scores were computed for the maximal span and for the total number of sequences they correctly repeated. A two-tailed $t$-test was conducted to compare the groups. The resulting effect was statistically significant for the maximal span, $t(55) = -2.5$, $p = .017$, the Cohen's d calculated on summary statistics is of $-.74$, critical $d = \pm .62$. But for the total number of sequences the test did not reach statistical significance ($t(55) = -1.8$, $p = 0.076$), actual $d = -0.31$ and critical $d$ again $\pm.62$. Thus, given the limited sample size, a result must be of rather large magnitude to be associated with $p < 0.05$, while medium to small effect sizes will simply fail to reach significance. This case makes it obvious that a failure to find a significant result does not imply absence of an effect, while a statistically significant effect may represent an overestimation. } Non-significant effects are likely, even if there is a true effect of interest. If a significant effect is observed, but most effects in a field are smaller than a study can detect, it is probable the effect size is overestimated  [@hedges1984estimation], and this should be taken into account when interpreting the effect size. 
\textcolor{red}{Beyond critical effect size values, researchers should reason before conducting a study on what conclusions can be drawn, and for those cases where sample size constrains will make it impossible to find an effect because it is likely smaller than the critical effect size value, alternative options for data collection, such as multi-lab studies (\cite{AAA}), should be explored.}

## Large Sample Sizes and Meta-analysis

When sample sizes are very large, the critical effect size values will make it clear that trivially small effect sizes will be statistically significant. Reporting critical effect size values will focus the attention of researchers on the difference between statistical significance and practical significance, and raises awareness of the importance to interpret the size of the effect. This is especially important in correlational studies with large sample sizes in psychology, where systematic but uncontrolled sources of variability may lead to small but theoretically meaningless non-zero effects, a phenomenon referred to as the 'crud factor' [@orben_crud_2020].

Imagine researchers who gain access to a very large archival dataset ($n = 5,000$) and decide to explore bivariate correlations between variables. With such a large sample size, very small correlations will reach statistical significance, and the critical effect size values for a two-sided test are $r = \pm 0.03$. It is worth noting that such small effects may potentially reflect just minor artifacts [@wilson2020science], such as subtle experimenter effects or slight non-independence among observations, even in meticulously designed studies.

The second real-world instance involves the study by @kramer2014experimental, which explored the impact of emotional content on Facebook users' experiences. With a notably large sample size of $n = 689.003$, the researchers observed a statistically significant increase in the number of negative words typed by participants when positive posts were reduced in their timeline, $t(310,044) = −5.63$, $p < 0.001$ (non-directional test). Based on this and other results from the study, the authors conclude that \textcolor{red}{"the emotions expressed by friends, via online social networks, influence our own moods, constituting, to our knowledge, the first experimental evidence for massive-scale emotional contagion via social networks".} The critical values for Cohen's $d$ in this experimental design are $-0.006$ and $+0.006$. It should have been clear from the outset that the researchers should have considered the effect size beyond merely testing for the statistical significance. The critical effect size should prompt reflection on which effect sizes are truly meaningful in this context, and whether the observed effect size ($d = 0.02$) is theoretically or practically interesting.

A similar scenario may arise in meta-analysis. Despite potential loss of precision due to substantial heterogeneity across effect sizes in different studies, meta-analyses typically synthesize a large amount of evidence, and as a consequence, even very small average effect sizes can reach significance. While the focus of meta-analysis is generally on estimating effect sizes with uncertainty, statistical significance is routinely reported and interpreted. Signaling the critical effect size values beforehand can serve as a clear warning that statistically significant results should not automatically be interpreted as practically significant, thus urging caution when interpreting the results.

@@

\bfilippo In this section, we provide guidance and formulas for computing standardized critical effect size with examples for frequently encountered effect sizes including Standardized Mean Differences (Cohen's $d$), correlations (Pearson's $r$), and raw and standardized coefficients in linear models. The same equations can be used to calculated unstandardized effect sizes that can be preferred in some situations (e.g., linear models). \efilippo These formulas have been incorporated into R functions of the package `criticalESvalue` accessible at: [https://github.com/psicostat/criticalESvalue](https://github.com/psicostat/criticalESvalue), and are elaborated upon in the subsequent section.

@@

Where $b$ is the unstandardized effect size that depends on the type of test. \bfilippo For example, in the two sample $t$-test the numerator is the difference between the two sample means while in the one-sample case is the difference between the sample mean and the population value. The denominator is the standard error of the numerator that depends on the sample size and the samples variances. \efilippo

\bfilippo
Equation \@ref(eq:eq-gen-d) describe a general formulation of a standardized effect size measure. Where $b$ is the unstandardized effect size (e.g., the difference between two means) and $s$ is the standardization term \efilippo. For example, in the two-samples case, $s$ is the pooled standard deviation between the two samples, while in the paired samples case, $s$ is the standard deviation of the differences for paired samples.

@@

@@
